---
title: "CLR_linearmodel"
output: html_document
---

```{r}
library(phyloseq)
library(vegan)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(ggplot2)
library(tidyverse)
library(ggh4x)
library(lme4)
library(lmerTest)
library(broom.mixed)
```

```{r}
IBD_path <- "../data/D003093_Colitis,_Ulcerative/species/phyloseq.RData"

ps_IBD <- get(load(IBD_path))

ps_IBD_rel <- transform_sample_counts(ps_IBD, function(x) x / sum(x))
otu_IBD_rel <- as(otu_table(ps_IBD_rel), "matrix") %>% t() %>% as.data.frame()
meta_IBD_rel <- as(sample_data(ps_IBD_rel), "matrix") %>% as.data.frame()
tax_IBD_rel <- as(tax_table(ps_IBD_rel), "matrix") %>% as.data.frame()  
```

# Compositional analysis with CLR transformation

Details: [9. Compositional Analysis with CLR Transformation][wiki-clr]. Transform counts with CLR before fitting the adjusted age model on a compositional scale.

```{r}
ps_clr <- transform_sample_counts(ps, function(x) {
  x <- x + 1e-9
  log(x) - mean(log(x))
})
otu_clr <- as(otu_table(ps_clr), "matrix") %>% t() %>% as.data.frame()
meta_clr <- as(sample_data(ps_clr), "matrix") %>% as.data.frame()
tax_clr <- as(tax_table(ps_clr), "matrix") %>% as.data.frame()
```
The hidden chunk below reshapes CLR values into the long table `dat_clr`.

```{r, include=FALSE}
dat_clr <- otu_clr %>%
  rownames_to_column("sample_id") %>%
  pivot_longer(-sample_id, names_to = "taxon", values_to = "clr") %>%
  left_join(meta_clr %>% as("data.frame") %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  left_join(tax_clr %>% rownames_to_column("taxon"), by = "taxon")
```
Fit one adjusted linear model per taxon using CLR values as the response.

```{r}
clr_age_adjusted_status_results <- dat_clr %>%
  filter(body_site == "gut") %>%
  mutate(
    age = factor(age, levels = c("infant", "adult")),
    status = factor(status, levels = c("healthy", "sick"))
  ) %>%
  group_by(taxon, genus, species) %>%
  group_modify(~ {
    fit <- lm(
      clr ~ age + status,
      data = .x
    )

    broom::tidy(fit)
  }) %>%
  ungroup() %>%
  filter(term == "ageadult") %>%
  mutate(q_value = p.adjust(p.value, method = "BH"))
```
Positive CLR estimates indicate adult enrichment relative to the community geometric mean; negative estimates indicate infant enrichment. The next table lists infant-enriched taxa.

```{r}
clr_infant_enriched <- clr_age_adjusted_status_results %>%
  select(taxon, genus, species, estimate, std.error, statistic, p.value, q_value) %>%
  arrange(estimate) %>%
  head(10)

clr_infant_enriched
```
Inspect the top CLR infant-enriched taxon on the CLR scale.

```{r}
dat_clr %>%
  filter(taxon == clr_infant_enriched %>% slice(1) %>% pull(taxon)) %>%
  filter(body_site == "gut") %>%
  ggplot(aes(x = age, y = clr, fill = age)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15) +
    theme_bw()
```
The y-axis is CLR, not relative abundance; compare direction rather than raw magnitude with earlier plots.

The next table lists CLR adult-enriched taxa.

```{r}
clr_adult_enriched <- clr_age_adjusted_status_results %>%
  select(taxon, genus, species, estimate, std.error, statistic, p.value, q_value) %>%
  arrange(-estimate) %>%
  head(10)

clr_adult_enriched
```
Inspect the top CLR adult-enriched taxon on the CLR scale.

```{r}
dat_clr %>%
  filter(taxon == clr_adult_enriched %>% slice(1) %>% pull(taxon)) %>%
  filter(body_site == "gut") %>%
  ggplot(aes(x = age, y = clr, fill = age)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15) +
    theme_bw()
```
Check whether the group separation matches the CLR estimate and q-value.

The CLR volcano plot summarises the compositional model results.

```{r}
# Volcano plot: CLR-based linear model
clr_age_adjusted_status_results %>%
  mutate(
    neg_log10_q = -log10(q_value),
    direction = case_when(
      q_value < 0.05 & estimate > 0 ~ "Adult-enriched",
      q_value < 0.05 & estimate < 0 ~ "Infant-enriched",
      TRUE ~ "Not significant"
    )
  ) %>%
  ggplot(aes(x = estimate, y = neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 2) +
    scale_color_manual(values = c("Adult-enriched" = "#E74C3C", "Infant-enriched" = "#3498DB", "Not significant" = "grey60")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    labs(x = "CLR estimate (adult vs infant)", y = "-log10(q-value)", color = NULL,
         title = "Compositional analysis (CLR): age effect in gut microbiome") +
    theme_bw()
```
Use this volcano plot as the compositionally aware reference analysis.

The next plot shows CLR effect size on the x-axis and orders taxa by q-value.

```{r}
# Bar chart of top significant taxa ordered by q-value (CLR)
bind_rows(
  clr_age_adjusted_status_results %>% filter(q_value < 0.05) %>% arrange(estimate) %>% head(10),
  clr_age_adjusted_status_results %>% filter(q_value < 0.05) %>% arrange(-estimate) %>% head(10)
) %>%
  distinct() %>%
  mutate(
    label = species,
    direction = ifelse(estimate > 0, "Adult-enriched", "Infant-enriched")
  ) %>%
  ggplot(aes(x = reorder(label, -q_value), y = estimate, fill = direction)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("Adult-enriched" = "#E74C3C", "Infant-enriched" = "#3498DB")) +
    labs(x = NULL, y = "CLR estimate (adult vs infant)", fill = NULL,
         title = "Top differentially abundant taxa (CLR)") +
    theme_bw()
```
Taxa that change strongly under CLR are important interpretation cases.


