---
title: "DMT2_full_analysis"
output: 
  html_document:
    toc: true
    toc_float: true

---

```{r}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
library(phyloseq)
library(vegan)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(ggplot2)
library(tidyverse)
library(ggh4x)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(microbiome)
```



```{r}

#load phyloseq object

DM_path <- "../data/D003924_Diabetes_Mellitus,_T/species/phyloseq.RData"
ps_dm <- get(load(DM_path))  # rename directly to ps
```

```{r}
sample_variables(ps_dm)
sample_data(ps_dm)
rank_names(ps_dm)
```


#Initially, when looking at the trends in sample metadata, we can notice some things that are important to have in mind: 
#1: Disease.MESH.ID: D003924 = people with the diabetes disease, D006262 = healthy people
#2: The number of samples from people who recently used antibiotics is way lower than people who did not recently use antibiotics (this is an imbalance in the dataset)
#3: Only samples from USA contain BMI information
#4: the last two columns (Disease.name and QC.status) are not contributing with valuable information to the dataset, because the value is the same for all the samples for both columns.

```{r}

#Working with the metadata is easier with a dataframe structure
ps_dm_df <- sample_data(ps_dm) %>% data.frame()


table(ps_dm_df$Sex)
table(ps_dm_df$Country)
table(ps_dm_df$Recent.antibiotics.use)
table(ps_dm_df$Host.age)
table(ps_dm_df$Disease.MESH.ID)
table(ps_dm_df$Disease.name)
table(ps_dm_df$BMI)
table(ps_dm_df$Experiment.type)
table(ps_dm_df$Project.ID)

```
#when summarizing each column, we get more information about the distribution of feature values. 

#The values for age and BMI have decimals, which is a bit tedious to work with, so I might have to group/floor the values before doing further analysis. Columns to remove (zero variance): Experiment.type, Disease.name, QC.status

#Additionally, there are 19 missing values for Sex and 54 missing values for recent antibiotics use.
#I didn't summarize Run.ID because it looks like a key for the different samples, so I would get 432 different values.
#Project.ID corresponds to the region in the world the studies come from. PRJNA422434 = China, PRJNA299502 = USA, PRJEB1786 = Europe (Sweden is the majority, and 1-2 samples from some other european countries)

#In conclusion: the availability of the different metadata measures is very much dependent on the individual study and thereby also the world region.

#Next step: Filtering, binning, etc.


```{r}


#Setting na's in the Sex column to "Unknown": code coming soon


#Making a column with interpretable binary values instead of disease MeSH ID's
sample_data(ps_dm_df)$Disease <- ifelse(
  sample_data(ps_dm_df)$Disease.MESH.ID == "D003924",
  "DMT2",
  "No_DMT2"
) %>% factor()


#Categorizing the geographical regions 
sample_data(ps_dm_df)$Study_region <- case_when(
  sample_data(ps_dm_df)$Project.ID ==
    "PRJEB1786" ~ "Europe",

  sample_data(ps_dm_df)$Project.ID ==
    "PRJNA299502" ~ "USA",

  sample_data(ps_dm_df)$Project.ID == 
    "PRJNA422434" ~ "China",

  TRUE ~ "Other"
)


#splitting age into age groups
sample_data(ps)$AgeGroup <- cut(
  sample_data(ps)$Age,
  breaks = c(10,30,50,70,90),
  labels = c("Young","Middle","Older","Elderly")
)




```




