---
title: "DMT2_full_analysis"
output: 
  html_document:
    toc: true
    toc_float: true

---

```{r}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
library(phyloseq)
library(vegan)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(ggplot2)
library(tidyverse)
library(ggh4x)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(microbiome)
```



```{r}

#load phyloseq object

DM_path <- "../data/D003924_Diabetes_Mellitus,_T/species/phyloseq.RData"
ps_dm <- get(load(DM_path))  # rename directly to ps
```

```{r}
sample_variables(ps_dm)
sample_data(ps_dm)
```


#Initially, when looking at the trends in sample metadata, we can notice some things that are important to have in mind: 
#1: Disease.MESH.ID: D003924 = people with the diabetes disease, D006262 = healthy people
#2: The number of samples from people who recently used antibiotics is way lower than people who did not recently use antibiotics (this is an imbalance in the dataset)
#3: Only samples from USA contain BMI information
#4: the last two columns (Disease.name and QC.status) are not contributing with valuable information to the dataset, because the value is the same for all the samples for both columns.

```{r}

#Working with the metadata is easier with a dataframe structure
ps_dm_df <- sample_data(ps_dm) %>% data.frame()


table(ps_dm_df$Sex)
table(ps_dm_df$Country)
table(ps_dm_df$Recent.antibiotics.use)
table(ps_dm_df$Host.age)
table(ps_dm_df$Disease.MESH.ID)
table(ps_dm_df$Disease.name)
table(ps_dm_df$BMI)
table(ps_dm_df$Experiment.type)
table(ps_dm_df$Project.ID)

```
#when summarizing each column, we get more information about the distribution of feature values. 
#The values for age and BMI have decimals, which is a bit tedious to work with, so I might have to group/floor the values before doing further analysis. Columns to remove: Experiment.type, Disease.name, QC.status


```{r}

```




