---
title: "Gut Microbiome Analysis: Diabetes and Geography"
output: html_document
---

# Overview

This document addresses three research questions:

1. **Global diabetes effect** — Is there any difference in the gut microbiome of diabetics vs. non-diabetics across the whole dataset?
2. **China-specific diabetes effect** — Is there any difference in the gut microbiome of diabetics vs. non-diabetics within China?
3. **East vs. West in diabetics** — Is there any geographical difference in the gut microbiome of diabetics across Western and Eastern lifestyles?

---

# Setup

## Libraries
```{r libraries, message=FALSE, warning=FALSE}
library(phyloseq)
library(vegan)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(ggplot2)
library(ape)
library(tidyverse)
library(ggh4x)
library(lme4)
library(lmerTest)
library(broom.mixed)
```

## Load Data
```{r load-data}
path <- "../data/D003924_Diabetes_Mellitus,_T/genus/phyloseq.RData"
ps <- get(load(path))

meta <- as(sample_data(ps), "matrix") %>% as.data.frame()
otu  <- as(otu_table(ps),   "matrix") %>% as.data.frame()
tax  <- as(tax_table(ps),   "matrix") %>% as.data.frame()
tree <- phy_tree(ps)
```

## Data Overview
```{r data-overview}
table(meta$Sex,                    useNA = "always")
table(meta$Country,                useNA = "always")
table(meta$Disease.MESH.ID,        useNA = "always")
table(meta$Recent.antibiotics.use, useNA = "always")
table(meta$Country, meta$Disease.MESH.ID, useNA = "always")

cat("Max library size:", max(sample_sums(ps)), "\n")
cat("NAs – Host.age:", sum(is.na(meta$Host.age)), "\n")
cat("NAs – BMI:",      sum(is.na(meta$BMI)),      "\n")

# Confirm data are relative abundances (all values ≤ 1)
cat("Values > 1 (should be 0 for rel. abundance):", sum(otu_table(ps) > 1), "\n")
```


## Prepare Phyloseq Objects and Subsets

Country labels are collapsed into West/East (China), and three analysis-specific
phyloseq objects are created.

```{r prepare-subsets}
# ── Recode western countries ──────────────────────────────────────────────────
europe <- c("Denmark", "Estonia", "Finland", "Germany",
             "Hungary", "Sweden", "Yugoslavia")

sample_data(ps)$Country <- as.character(sample_data(ps)$Country)
sample_data(ps)$Country[sample_data(ps)$Country %in% europe] <- "Europe"
sample_data(ps)$Country <- factor(sample_data(ps)$Country)

# ── Analysis subsets ──────────────────────────────────────────────────────────
# Analysis 1: whole dataset (diabetics D003924 vs healthy D006262)
ps_global <- ps

# Analysis 2: China only (diabetics vs healthy)
ps_china  <- subset_samples(ps, Country == "China")

# Analysis 3: diabetics only (West vs China)
ps_we     <- subset_samples(ps, Disease.MESH.ID == "D003924" & Sex == "Female" &  Country %in% c("Europe", "China"))
```

## Transformed Phyloseq Objects
```{r transforms}
make_transforms <- function(ps) {
  list(
    rel = transform_sample_counts(ps, function(x) x / sum(x)),
    pa  = transform_sample_counts(ps, function(x) ifelse(x > 0, 1, 0)),
    clr = transform_sample_counts(ps, function(x) { x <- x + 1e-9; log(x) - mean(log(x)) })
  )
}

global_t <- make_transforms(ps_global)
china_t  <- make_transforms(ps_china)
we_t     <- make_transforms(ps_we)
```

### Data 
```{r}
ps_rel <- transform_sample_counts(ps, function(x) x / sum(x))
otu_rel <- as(otu_table(ps_rel), "matrix") %>% t() %>% as.data.frame()
meta_rel <- as(sample_data(ps_rel), "matrix") %>% as.data.frame()
tax_rel <- as(tax_table(ps_rel), "matrix") %>% as.data.frame()  
ps_genus <- tax_glom(ps_rel, taxrank = "Species")

dat_rel <- otu_rel %>%
  rownames_to_column("sample_id") %>%
  pivot_longer(-sample_id, names_to = "taxon", values_to = "rel_abund") %>%
  left_join(meta_rel %>% as("data.frame") %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  left_join(tax_rel %>% rownames_to_column("taxon"), by = "taxon")

one_taxon <- dat_rel$taxon[3]


# Comparison between body sites
dat_rel %>%
  filter(taxon == one_taxon) %>%
  ggplot(aes(x = Disease.MESH.ID, y = rel_abund, fill = Disease.MESH.ID)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15) +
    theme_bw()

```

```{r}
dat_rel %>%
  filter(Disease.MESH.ID == "D003924") %>%
  group_by(taxon) %>%
  summarise(prop_zero = mean(rel_abund == 0), .groups = "drop") %>%
  ggplot(aes(x = prop_zero)) +
    geom_histogram(binwidth = 0.05, fill = "steelblue", color = "white") +
    labs(x = "Proportion of samples where relative abundance = 0",
         y = "Number of taxa",
         title = "Zero-inflation: most taxa are absent from most samples") +
    theme_bw()
```
```{r}
set.seed(42)
example_taxa <- dat_rel %>%
  filter(Disease.MESH.ID == "D006262") %>%
  group_by(taxon, Genus, Species) %>%
  summarise(prop_nonzero = mean(rel_abund > 0), .groups = "drop") %>%
  filter(prop_nonzero > 0.4, prop_nonzero < 0.9) %>%
  slice_sample(n = 6) %>%
  pull(taxon)

dat_rel %>%
  filter(Disease.MESH.ID == "D006262", taxon %in% example_taxa, rel_abund > 0) %>%
  mutate(label = Species) %>%
  ggplot(aes(x = rel_abund)) +
    geom_histogram(bins = 20, fill = "steelblue", color = "white") +
    facet_wrap(~ label, scales = "free") +
    labs(x = "Relative abundance (non-zero values only)",
         y = "Count",
         title = "Right-skewed distributions in gut taxa") +
    theme_bw()
```
```{r}
dat_rel %>%
  filter(Disease.MESH.ID == "D006262", taxon %in% example_taxa) %>%
  mutate(label = Species) %>%
  ggplot(aes(sample = rel_abund)) +
    stat_qq(size = 0.8, alpha = 0.6) +
    stat_qq_line(color = "red", linetype = "dashed") +
    facet_wrap(~ label, scales = "free") +
    labs(x = "Theoretical quantiles (normal)",
         y = "Sample quantiles",
         title = "Q–Q plots: departures from normality in gut taxa") +
    theme_bw()
```

## Helper Functions
```{r helpers}
# Random subsample of samples from a phyloseq object
sample_ps <- function(ps, FUN = sample, ...) {
  prune_samples(FUN(sample_names(ps), ...), ps)
}

# Alpha-diversity data frame (Shannon + Observed) joined to metadata
make_alpha_df <- function(ps, ps_pa) {
  shannon  <- estimate_richness(ps,    measures = "Shannon")  %>% rownames_to_column("sample_id")
  observed <- estimate_richness(ps_pa, measures = "Observed") %>% rownames_to_column("sample_id")
  meta_df  <- sample_data(ps) %>% data.frame() %>% rownames_to_column("sample_id")

  inner_join(shannon, observed, by = "sample_id") %>%
    inner_join(meta_df, by = "sample_id")
}

# Kruskal-Wallis over multiple grouping variables
run_kruskal <- function(data, response_var, group_vars) {
  map_dfr(group_vars, function(gv) {
    res <- kruskal.test(as.formula(paste(response_var, "~", gv)), data = data)
    data.frame(Variable = gv, Chi_squared = res$statistic,
               df = res$parameter, P_value = res$p.value)
  })
}

# Pairwise Wilcoxon over multiple grouping variables
run_pairwise_wilcoxon <- function(data, response_var, group_vars) {
  map_dfr(group_vars, function(gv) {
    res <- pairwise.wilcox.test(data[[response_var]], data[[gv]],
                                p.adjust.method = "bonferroni")
    as.data.frame(res$p.value) %>%
      rownames_to_column("Group1") %>%
      pivot_longer(-Group1, names_to = "Group2", values_to = "P_value_adj") %>%
      drop_na(P_value_adj) %>%
      mutate(Variable = gv, Comparison = paste(Group1, "vs", Group2)) %>%
      select(Variable, Comparison, P_value_adj, Group1, Group2)
  })
}

# CLR linear model across all taxa, returning results for one term
run_clr_lm <- function(ps_clr, formula_str, term_filter) {
  otu_clr  <- as(otu_table(ps_clr),  "matrix") %>% t() %>% as.data.frame()
  meta_clr <- as(sample_data(ps_clr), "matrix") %>% as.data.frame()
  tax_clr  <- as(tax_table(ps_clr),  "matrix") %>% as.data.frame()

  otu_clr %>%
    rownames_to_column("sample_id") %>%
    pivot_longer(-sample_id, names_to = "taxon", values_to = "clr") %>%
    left_join(meta_clr %>% rownames_to_column("sample_id"), by = "sample_id") %>%
    left_join(tax_clr  %>% rownames_to_column("taxon"),     by = "taxon") %>%
    mutate(
      sex    = factor(Sex,            levels = c("Female", "Male")),
      status = factor(Disease.MESH.ID, levels = c("D006262", "D003924")),
      country= factor(Country,         levels = c("China", "Europe", "United States of America")),
      age    = as.numeric(Host.age),
      bmi    = as.numeric(BMI),
      depth  = as.numeric(Nr..reads.sequenced)
    ) %>%
    group_by(taxon, Genus, Species) %>%
    group_modify(~ broom::tidy(lm(as.formula(formula_str), data = .x))) %>%
    filter(term == term_filter) %>%
    ungroup() %>%
    mutate(q_value = p.adjust(p.value, method = "BH"))
}

# Volcano plot from CLR results
plot_volcano <- function(clr_res, x_label, title,
                         pos_label = "Enriched (+)", neg_label = "Enriched (-)") {
  clr_res %>%
    mutate(
      neg_log10_q = -log10(q_value),
      direction   = case_when(
        q_value < 0.05 & estimate > 0 ~ pos_label,
        q_value < 0.05 & estimate < 0 ~ neg_label,
        TRUE ~ "Not significant"
      )
    ) %>%
    ggplot(aes(x = estimate, y = neg_log10_q, color = direction)) +
      geom_point(alpha = 0.6, size = 2) +
      scale_color_manual(values = c(setNames("#E74C3C", pos_label),
                                    setNames("#3498DB", neg_label),
                                    "Not significant" = "grey60")) +
      geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
      geom_vline(xintercept = 0,            linetype = "dashed", color = "grey40") +
      labs(x = x_label, y = "-log10(q-value)", color = NULL, title = title) +
      theme_bw()
}

# Top-taxa bar chart from CLR results
plot_top_taxa <- function(clr_res, n = 10, x_label, title,
                          pos_label = "Enriched (+)", neg_label = "Enriched (-)") {
  bind_rows(
    clr_res %>% filter(q_value < 0.05) %>% arrange(estimate)  %>% head(n),
    clr_res %>% filter(q_value < 0.05) %>% arrange(-estimate) %>% head(n)
  ) %>%
    distinct() %>%
    mutate(direction = ifelse(estimate > 0, pos_label, neg_label)) %>%
    ggplot(aes(x = reorder(Species, -q_value), y = estimate, fill = direction)) +
      geom_col() +
      coord_flip() +
      scale_fill_manual(values = c(setNames("#E74C3C", pos_label),
                                   setNames("#3498DB", neg_label))) +
      labs(x = NULL, y = x_label, fill = NULL, title = title) +
      theme_bw()
}

# RDA on CLR matrix: fit model, run permutation ANOVA, return rda object + clean matrices
run_rda <- function(ps_clr, formula_str, covariates) {
  otu_clr  <- as(otu_table(ps_clr),   "matrix") %>% t()
  meta_clr <- as(sample_data(ps_clr), "matrix") %>% as.data.frame() %>%
  mutate(
        across(c(Host.age, BMI, Nr..reads.sequenced), as.numeric),
        # Explicitly set factor levels to match run_clr_lm
        Sex             = factor(Sex,             levels = c("Female", "Male")),
        Disease.MESH.ID = factor(Disease.MESH.ID, levels = c("D006262", "D003924")),
        Country         = factor(Country,         levels = c("China", "Europe", "United States of America"))
      )

  na_idx       <- apply(meta_clr[, covariates, drop = FALSE], 1, anyNA)
  clr_clean    <- otu_clr[!na_idx, ]
  meta_clean   <- meta_clr[!na_idx, ]

  cat("Samples removed due to NAs:", sum(na_idx), "\n")
  cat("Samples retained:", nrow(clr_clean), "\n")

  rda_fit <- rda(as.formula(paste("clr_clean ~", formula_str)), data = meta_clean)

  list(rda = rda_fit, clr_mat = clr_clean, meta = meta_clean)
}

# RDA biplot coloured by one variable, shaped by another (both optional)
plot_rda_biplot <- function(rda_obj, meta_df,
                            color_var, shape_var = NULL,
                            color_values, shape_values = NULL,
                            title = "RDA biplot") {
  eig  <- rda_obj$CCA$eig
  pct1 <- round(eig[1] / rda_obj$tot.chi * 100, 1)
  pct2 <- round(eig[2] / rda_obj$tot.chi * 100, 1)

  site_scores <- scores(rda_obj, display = "sites") %>%
    as.data.frame() %>%
    rownames_to_column("sample_id") %>%
    left_join(meta_df %>% rownames_to_column("sample_id"), by = "sample_id")

  bp_scores <- scores(rda_obj, display = "bp") %>%
    as.data.frame() %>%
    rownames_to_column("variable")

  site_range  <- max(abs(c(site_scores$RDA1, site_scores$RDA2)))
  arrow_scale <- site_range * 0.8 / max(sqrt(bp_scores$RDA1^2 + bp_scores$RDA2^2))
  bp_scores   <- bp_scores %>% mutate(RDA1 = RDA1 * arrow_scale, RDA2 = RDA2 * arrow_scale)

  aes_map <- if (!is.null(shape_var)) {
    aes(x = RDA1, y = RDA2,
        color = .data[[color_var]], shape = .data[[shape_var]])
  } else {
    aes(x = RDA1, y = RDA2, color = .data[[color_var]])
  }

  
  
  p <- ggplot() +
    geom_point(data = site_scores, mapping = aes_map, size = 3, alpha = 0.85) +
  geom_segment(data = top_species_scores,
               aes(x = 0, y = 0, xend = RDA1, yend = RDA2),
               arrow = arrow(length = unit(0.15, "cm")),
               color = "grey70", linewidth = 0.5) +
  geom_text(data = top_species_scores,
            aes(x = RDA1 * 1.15, y = RDA2 * 1.15, label = label),
            size = 2.5, color = "grey50") +
    geom_segment(data = bp_scores,
                 aes(x = 0, y = 0, xend = RDA1, yend = RDA2),
                 arrow = arrow(length = unit(0.25, "cm")),
                 color = "black", linewidth = 1) +
    geom_text(data = bp_scores,
              aes(x = RDA1 * 1.15, y = RDA2 * 1.15, label = variable),
              size = 3.5, fontface = "bold") +
    scale_color_manual(values = color_values) +
    labs(x = paste0("RDA1 (", pct1, "% of total variance)"),
         y = paste0("RDA2 (", pct2, "% of total variance)"),
         title = title) +
    theme_bw()

  if (!is.null(shape_var) && !is.null(shape_values)) {
    p <- p + scale_shape_manual(values = shape_values)
  }
  p
}

# Join RDA taxon loadings with CLR univariate estimates and flag agreement
compare_rda_clr <- function(rda_obj, clr_res, tax_df, rda1_sign = 1) {
  taxon_scores <- scores(rda_obj, display = "species")[, 1] * rda1_sign

  tibble(taxon = names(taxon_scores), rda1_loading = taxon_scores) %>%
    left_join(tax_df %>% rownames_to_column("taxon"), by = "taxon") %>%
    left_join(
      clr_res %>% select(taxon, Genus, Species,
                         clr_estimate = estimate, clr_q = q_value),
      by = "taxon"
    ) %>%
    mutate(
      direction_agreement = sign(rda1_loading) == sign(clr_estimate),
      clr_result = case_when(
        clr_q < 0.05 & direction_agreement  ~ "CLR significant, same direction",
        clr_q < 0.05 & !direction_agreement ~ "CLR significant, opposite direction",
        TRUE                                ~ "CLR not significant"
      )
    ) %>%
    arrange(desc(abs(rda1_loading)))
}

# Scatter plot: RDA1 loading vs CLR estimate, coloured by agreement
plot_rda_clr_comparison <- function(comparison_df, x_label, y_label, title) {
  comparison_df %>%
    ggplot(aes(x = clr_estimate, y = rda1_loading, color = clr_result)) +
      geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
      geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
      geom_point(alpha = 0.65, size = 2) +
      scale_color_manual(values = c(
        "CLR significant, same direction"      = "#2E8B57",
        "CLR significant, opposite direction"  = "#E67E22",
        "CLR not significant"                  = "grey70"
      )) +
      labs(x = x_label, y = y_label, color = NULL, title = title) +
      theme_bw()
}
```

---

# Analysis 1: Global — Diabetics vs. Non-Diabetics (Whole Dataset)

**Question:** Is there a difference in the gut microbiome between diabetics (D003924)
and non-diabetics (D006262) across all countries?

## Visualisation

### Barplot (relative abundance, n = 50 sample)
```{r a1-barplot, fig.width=10, fig.height=5}
ps_genus_global <- tax_glom(global_t$rel, taxrank = "Genus")
ps_genus_sampled <- sample_ps(ps_genus_global, size = 50)

plot_bar(ps_genus_sampled, fill = "Genus") +
  facet_nested(~ Disease.MESH.ID, scales = "free_x") +
  labs(title = "Relative abundance by disease status (global, n = 50)") +
  theme(axis.text.x = element_blank())
```
### Heat map
```{r}
plot_heatmap(ps_genus_sampled, taxa.label = "Genus", sample.order = 'Disease.MESH.ID')
```



## Alpha Diversity
```{r a1-alpha}
alpha_global <- make_alpha_df(ps_global, global_t$pa)

plot_richness(ps_global, measures = "Shannon", color = "Disease.MESH.ID") +
  labs(title = "Shannon diversity — global") +
  theme_bw()


plot_richness(global_t$pa, measures = "Observed", color = "Disease.MESH.ID") +
  labs(title = "Observed diversity — global") +
  theme_bw()

run_kruskal(alpha_global, "Shannon", c("Disease.MESH.ID", "Country"))
run_pairwise_wilcoxon(alpha_global, "Shannon", c("Disease.MESH.ID","Country"))


```

### Sequencing depth vs. Shannon
Negative correlation might be due to the removal of singletons.
```{r a1-depth-shannon}
alpha_global$reads <- sample_sums(ps_global)[alpha_global$sample_id]

ggplot(alpha_global, aes(x = reads, y = Shannon)) +
  geom_point() + geom_smooth(method = "lm") +
  labs(title = "Sequencing depth vs. Shannon — global") +
  theme_bw()

cor.test(alpha_global$reads, alpha_global$Shannon, method = "spearman")
```

## Beta Diversity
### Calculating the Bray-Curtis distances and the Jaccard distances
```{r}
bray_dist <- distance(
  ps_global,
  method = "bray"
)

jaccard_dist <- distance(
  global_t$pa,
  method = "jaccard"
)
```

### PCoA — Bray-Curtis
```{r a1-pcoa, fig.width=8, fig.height=5}
bray_global <- distance(global_t$rel, method = "bray")
ord_global  <- ordinate(global_t$rel, method = "PCoA", distance = bray_global)

plot_ordination(global_t$rel, ord_global, type = "samples",
                color = "Disease.MESH.ID", shape = "Country") +
  scale_color_manual(values = c("D003924" = "#E74C3C", "D006262" = "#3498DB"),
                     labels = c("Diabetic", "Healthy")) +
  labs(title = "PCoA (Bray-Curtis) — global dataset",
       color = "Status") +
  theme_bw()

plot_scree(ord_global)
```
### NMDS based on Bray-Curtis distance

NMDS is another commonly used ordination method in ecology and microbiome analysis. Unlike PCoA, it focuses on preserving the rank order of distances rather than the distances themselves.

```{r nmds-bray-global, fig.width=8, fig.height=5}
global_nmds <- ordinate(
  global_t$rel,
  method = "NMDS",
  distance = "bray"
)

plot_ordination(
  global_t$rel,
  global_nmds,
  type = "samples",
  color = "Disease.MESH.ID",
  shape = "Country"
)
```

### PERMANOVA + Dispersion
```{r a1-permanova}
sample_df_global <- sample_data(global_t$rel) %>% data.frame()

adonis2(bray_global ~ Disease.MESH.ID,
        data = sample_df_global, permutations = 999)

# Dispersion check
disp_global <- betadisper(bray_global, sample_df_global$Disease.MESH.ID)
anova(disp_global)
plot(disp_global, main = "Dispersion — global by disease status")
```

## Compositional Analysis — CLR Linear Model

The model adjusts for sex, country, age, and sequencing depth; the term of interest
is `statusD003924` (diabetic relative to healthy).

### Running CLR
```{r a1-clr}
clr_global <- run_clr_lm(
  ps_clr       = global_t$clr,
  formula_str  = "clr ~ sex + country + status + age + depth", 
  term_filter  = "statusD003924"
)


global_healthy_enriched <- clr_global %>%
  select(taxon, Genus, Species, estimate, std.error, statistic, p.value, q_value) %>%
  arrange(estimate) %>%
  head(10)

global_sick_enriched <- clr_global %>%
  select(taxon, Genus, Species, estimate, std.error, statistic, p.value, q_value) %>%
  arrange(-estimate) %>%
  head(10)

global_healthy_enriched
global_sick_enriched
```

### Plots
```{r}
plot_volcano(clr_global,
             x_label   = "CLR estimate (diabetic vs healthy)",
             title     = "CLR volcano — global: diabetic vs healthy",
             pos_label = "Diabetic-enriched",
             neg_label = "Healthy-enriched")

plot_top_taxa(clr_global,
              x_label   = "CLR estimate (diabetic vs healthy)",
              title     = "Top differentially abundant taxa — global",
              pos_label = "Diabetic-enriched",
              neg_label = "Healthy-enriched")
```

## Constrained Ordination — RDA

The CLR matrix is constrained by disease status, country, sex, age, and sequencing
depth. This tests which predictors explain significant variation in the full
community composition simultaneously.

```{r a1-rda}
rda_global_out <- run_rda(
  ps_clr      = global_t$clr,
  formula_str = "Disease.MESH.ID + Country + Sex + as.numeric(Host.age) + as.numeric(Nr..reads.sequenced)",
  covariates  = c("Disease.MESH.ID", "Country", "Sex", "Host.age", "Nr..reads.sequenced")
)

# Permutation ANOVA — overall model and per-term
anova(rda_global_out$rda, permutations = 999)
anova(rda_global_out$rda, by = "term", permutations = 999)
```

```{r a1-rda-biplot, fig.width=9, fig.height=6}
plot_rda_biplot(
  rda_obj      = rda_global_out$rda,
  meta_df      = rda_global_out$meta,
  color_var    = "Disease.MESH.ID",
  shape_var    = "Country",
  color_values = c("D003924" = "#E74C3C", "D006262" = "#3498DB"),
  shape_values = c("China" = 16, "West" = 17),
  title        = "RDA biplot — global: disease status and country"
)
```

### RDA vs CLR comparison
```{r a1-rda-clr-compare, fig.width=8, fig.height=5}
# Sign of RDA1 is arbitrary; orient so that the diabetic group has positive RDA1
rda1_scores_global <- scores(rda_global_out$rda, display = "sites")[, "RDA1"]
rda1_sign_global <- ifelse(
  mean(rda1_scores_global[rda_global_out$meta$Disease.MESH.ID == "D003924"], na.rm = TRUE) >=
  mean(rda1_scores_global[rda_global_out$meta$Disease.MESH.ID == "D006262"], na.rm = TRUE),
  1, -1
)

comp_global <- compare_rda_clr(
  rda_obj  = rda_global_out$rda,
  clr_res  = clr_global,
  tax_df   = as(tax_table(global_t$clr), "matrix") %>% as.data.frame(),
  rda1_sign = rda1_sign_global
)

plot_rda_clr_comparison(
  comp_global,
  x_label = "CLR estimate (diabetic vs healthy)",
  y_label = "RDA1 loading (positive = diabetic side)",
  title   = "RDA loadings vs CLR estimates — global"
)
```

```{r a1-rda-clr-table}
comp_global %>%
  arrange(desc(abs(rda1_loading))) %>%
  select(Genus.x, Species.x, rda1_loading, clr_estimate, clr_q, direction_agreement) %>%
  head(20)
```

---

# Analysis 2: China — Diabetics vs. Non-Diabetics

**Question:** Within China, is there a difference in the gut microbiome between
diabetics and non-diabetics?

## Visualisation

### Barplot (relative abundance, n = 50 sample)
```{r a2-barplot, fig.width=10, fig.height=5}
ps_genus_china  <- tax_glom(china_t$rel, taxrank = "Genus")
china_genus_sampled <- sample_ps(ps_genus_china, size = min(50, nsamples(ps_genus_china)))

plot_bar(china_genus_sampled, fill = "Genus") +
  facet_nested(~ Disease.MESH.ID, scales = "free_x") +
  labs(title = "Relative abundance by disease status (China)") +
  theme(axis.text.x = element_blank())
```


## Alpha Diversity
```{r a2-alpha}
alpha_china <- make_alpha_df(ps_china, china_t$pa)

plot_richness(ps_china, measures = "Shannon", color = "Disease.MESH.ID") +
  labs(title = "Shannon diversity — China") +
  theme_bw()

run_kruskal(alpha_china, "Shannon", "Disease.MESH.ID")
run_pairwise_wilcoxon(alpha_china, "Shannon", "Disease.MESH.ID")
```

### Sequencing depth vs. Shannon
```{r a2-depth-shannon}
alpha_china$reads <- sample_sums(ps_china)[alpha_china$sample_id]

ggplot(alpha_china, aes(x = reads, y = Shannon)) +
  geom_point() + geom_smooth(method = "lm") +
  labs(title = "Sequencing depth vs. Shannon — China") +
  theme_bw()

cor.test(alpha_china$reads, alpha_china$Shannon, method = "spearman")
```

## Beta Diversity — PERMANOVA
### Calculating the Bray-Curtis distances and the Jaccard distances
```{r}
bray_dist <- distance(
  ps_china,
  method = "bray"
)

jaccard_dist <- distance(
  china_t$pa,
  method = "jaccard"
)
```

### PCoA — Bray-Curtis
```{r a2-pcoa, fig.width=8, fig.height=5}
bray_china <- distance(china_t$rel, method = "bray")
ord_china  <- ordinate(china_t$rel, method = "PCoA", distance = bray_china)

plot_ordination(china_t$rel, ord_china, type = "samples",
                color = "Disease.MESH.ID") +
  scale_color_manual(values = c("D003924" = "#E74C3C", "D006262" = "#3498DB"),
                     labels = c("Diabetic", "Healthy")) +
  labs(title = "PCoA (Bray-Curtis) — China",
       color = "Status") +
  theme_bw()
```

### NMDS based on Bray-Curtis distance

NMDS is another commonly used ordination method in ecology and microbiome analysis. Unlike PCoA, it focuses on preserving the rank order of distances rather than the distances themselves.

```{r nmds-bray-china, fig.width=8, fig.height=5}
global_nmds <- ordinate(
  china_t$rel,
  method = "NMDS",
  distance = "bray"
)

plot_ordination(
  china_t$rel,
  global_nmds,
  type = "samples",
  color = "Disease.MESH.ID",
  shape = "Sex"
)
```

### PERMANOVA + Dispersion
```{r a2-permanova}
sample_df_china <- sample_data(china_t$rel) %>% data.frame()

adonis2(bray_china ~ Disease.MESH.ID,
        data = sample_df_china, permutations = 999)

disp_china <- betadisper(bray_china, sample_df_china$Disease.MESH.ID)
anova(disp_china)
plot(disp_china, main = "Dispersion — China by disease status")
```

## Compositional Analysis — CLR Linear Model

The model adjusts for sex, age, and sequencing depth (country is fixed to China,
so it is omitted). The term of interest is `statusD003924`.

### Running CLR + GLM
```{r a2-clr}
clr_china <- run_clr_lm(
  ps_clr      = china_t$clr,
  formula_str = "clr ~ sex + status + age + depth",
  term_filter = "statusD003924"
)


china_healthy_enriched <- clr_china %>%
  select(taxon, Genus, Species, estimate, std.error, statistic, p.value, q_value) %>%
  arrange(estimate) %>%
  head(10)

china_sick_enriched <- clr_china %>%
  select(taxon, Genus, Species, estimate, std.error, statistic, p.value, q_value) %>%
  arrange(-estimate) %>%
  head(10)

china_healthy_enriched
china_sick_enriched
```

### Plots
```{r plots}
plot_volcano(clr_china,
             x_label   = "CLR estimate (diabetic vs healthy, China)",
             title     = "CLR volcano — China: diabetic vs healthy",
             pos_label = "Diabetic-enriched",
             neg_label = "Healthy-enriched")

plot_top_taxa(clr_china,
              x_label   = "CLR estimate (diabetic vs healthy, China)",
              title     = "Top differentially abundant taxa — China",
              pos_label = "Diabetic-enriched",
              neg_label = "Healthy-enriched")
```

## Constrained Ordination — RDA

Within China, the CLR matrix is constrained by disease status, sex, age, and
sequencing depth (country is constant so it is omitted).

```{r a2-rda}
rda_china_out <- run_rda(
  ps_clr      = china_t$clr,
  formula_str = "Disease.MESH.ID + Sex + as.numeric(Host.age) + as.numeric(Nr..reads.sequenced)",
  covariates  = c("Disease.MESH.ID", "Sex", "Host.age", "Nr..reads.sequenced")
)

anova(rda_china_out$rda, permutations = 999)
anova(rda_china_out$rda, by = "term", permutations = 999)
```

```{r a2-rda-biplot, fig.width=9, fig.height=6}
plot_rda_biplot(
  rda_obj      = rda_china_out$rda,
  meta_df      = rda_china_out$meta,
  color_var    = "Disease.MESH.ID",
  shape_var    = "Sex",
  color_values = c("D003924" = "#E74C3C", "D006262" = "#3498DB"),
  shape_values = c("Female" = 16, "Male" = 17),
  title        = "RDA biplot — China: disease status"
)
```

### RDA vs CLR
```{r a2-rda-clr-compare, fig.width=8, fig.height=5}
# Sign of RDA1 is arbitrary; orient so that the diabetic group has positive RDA1
rda1_scores_china <- scores(rda_china_out$rda, display = "sites")[, "RDA1"]
rda1_sign_china <- ifelse(
  mean(rda1_scores_china[rda_china_out$meta$Disease.MESH.ID == "D003924"], na.rm = TRUE) >=
  mean(rda1_scores_china[rda_china_out$meta$Disease.MESH.ID == "D006262"], na.rm = TRUE),
  1, -1
)

comp_china <- compare_rda_clr(
  rda_obj  = rda_china_out$rda,
  clr_res  = clr_china,
  tax_df   = as(tax_table(global_t$clr), "matrix") %>% as.data.frame(),
  rda1_sign = rda1_sign_china
)

plot_rda_clr_comparison(
  comp_china,
  x_label = "CLR estimate (diabetic vs healthy)",
  y_label = "RDA1 loading (positive = diabetic side)",
  title   = "RDA loadings vs CLR estimates — global"
)
```

```{r a2-rda-clr-table}
comp_china %>%
  arrange(desc(abs(rda1_loading))) %>%
  select(Genus.x, Species.x, rda1_loading, clr_estimate, clr_q, direction_agreement) %>%
  head(5)
```

# Analysis 3: East vs. West — Geographical Differences in Diabetics

**Question:** Among diabetic individuals only, is there a geographical difference
in gut microbiome composition between Western and Chinese (Eastern) populations?

## Visualisation

### Barplot (relative abundance, n = 50 sample)
```{r a3-barplot, fig.width=10, fig.height=5}
ps_genus_we  <- tax_glom(we_t$rel, taxrank = "Genus")
we_genus_sampled <- sample_ps(ps_genus_we, size = min(50, nsamples(ps_genus_we)))

plot_bar(we_genus_sampled, fill = "Genus") +
  facet_nested(~ Country, scales = "free_x") +
  labs(title = "Relative abundance by geography (diabetics only)") +
  theme(axis.text.x = element_blank())
```


## Alpha Diversity
```{r a3-alpha}
alpha_we <- make_alpha_df(ps_we, we_t$pa)

plot_richness(ps_we, measures = "Shannon", color = "Country") +
  scale_color_manual(values = c("China" = "#E74C3C", "West" = "#3498DB")) +
  labs(title = "Shannon diversity — diabetics by geography") +
  theme_bw()

run_kruskal(alpha_we, "Shannon", "Country")
run_pairwise_wilcoxon(alpha_we, "Shannon", "Country")
```

### Sequencing depth vs. Shannon
```{r a3-depth-shannon}
alpha_we$reads <- sample_sums(ps_we)[alpha_we$sample_id]

ggplot(alpha_we, aes(x = reads, y = Shannon, color = Country)) +
  geom_point() + geom_smooth(method = "lm") +
  scale_color_manual(values = c("China" = "#E74C3C", "Europe" = "#3498DB")) +
  labs(title = "Sequencing depth vs. Shannon — diabetics by geography") +
  theme_bw()

cor.test(alpha_we$reads, alpha_we$Shannon, method = "spearman")
```

## Beta Diversity — PERMANOVA
### Calculating the Bray-Curtis distances and the Jaccard distances
```{r}
bray_dist <- distance(
  ps_we,
  method = "bray"
)

jaccard_dist <- distance(
  we_t$pa,
  method = "jaccard"
)
```

### PCoA — Bray-Curtis
```{r a3-pcoa, fig.width=8, fig.height=5}
bray_we <- distance(we_t$rel, method = "bray")
ord_we  <- ordinate(we_t$rel, method = "PCoA", distance = bray_we)

plot_ordination(we_t$rel, ord_we, type = "samples", color = "Country") +
  scale_color_manual(values = c("China" = "#E74C3C", "Europe" = "#3498DB")) +
  labs(title = "PCoA (Bray-Curtis) — diabetics: East vs. West") +
  theme_bw()
```

### NMDS based on Bray-Curtis distance

NMDS is another commonly used ordination method in ecology and microbiome analysis. Unlike PCoA, it focuses on preserving the rank order of distances rather than the distances themselves.

```{r nmds-bray-we, fig.width=8, fig.height=5}
we_nmds <- ordinate(
  we_t$rel,
  method = "NMDS",
  distance = "bray"
)

plot_ordination(
  we_t$rel,
  we_nmds,
  type = "samples",
  color = "Country",
  shape = "Sex"
)
```

### PERMANOVA + Dispersion
```{r a3-permanova}
sample_df_we <- sample_data(we_t$rel) %>% data.frame()

adonis2(bray_we ~ Country,
        data = sample_df_we, permutations = 999)

disp_we <- betadisper(bray_we, sample_df_we$Country)
anova(disp_we)
plot(disp_we, main = "Dispersion — diabetics by geography")
```

## Compositional Analysis — CLR Linear Model (Constrained Ordination — RDA)

The model adjusts for sex, age, BMI, and sequencing depth. The term of interest
is `countryWest` (West relative to China).

```{r a3-clr}
clr_we <- run_clr_lm(
  ps_clr      = we_t$clr,
  formula_str = "clr ~ country + age + depth",
  term_filter = "countryEurope"
)

we_china_enriched <- clr_we %>%
  select(taxon, Genus, Species, estimate, std.error, statistic, p.value, q_value) %>%
  arrange(estimate) %>%
  head(10)

we_europe_enriched <- clr_we %>%
  select(taxon, Genus, Species, estimate, std.error, statistic, p.value, q_value) %>%
  arrange(-estimate) %>%
  head(10)

we_china_enriched
we_europe_enriched
```

```{r}
plot_volcano(clr_we,
             x_label   = "CLR estimate (West vs China, diabetics)",
             title     = "CLR volcano — diabetics: West vs. China",
             pos_label = "West-enriched",
             neg_label = "China-enriched")

plot_top_taxa(clr_we,
              x_label   = "CLR estimate (West vs China, diabetics)",
              title     = "Top differentially abundant taxa — East vs. West (diabetics)",
              pos_label = "West-enriched",
              neg_label = "China-enriched")
```

### PCA of CLR Profiles
```{r a3-pca-clr, fig.width=8, fig.height=5}
otu_clr_we  <- as(otu_table(we_t$clr), "matrix") %>% t()
meta_clr_we <- as(sample_data(we_t$clr), "matrix") %>% as.data.frame()

pca_we  <- prcomp(otu_clr_we, center = TRUE, scale. = FALSE)
pca_var <- round(100 * pca_we$sdev^2 / sum(pca_we$sdev^2), 1)

as.data.frame(pca_we$x[, 1:2]) %>%
  rownames_to_column("sample_id") %>%
  left_join(meta_clr_we %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  ggplot(aes(x = PC1, y = PC2, color = Country)) +
    geom_point(size = 3, alpha = 0.85) +
    scale_color_manual(values = c("China" = "#E74C3C", "Europe" = "#3498DB")) +
    labs(x = paste0("PC1 (", pca_var[1], "%)"),
         y = paste0("PC2 (", pca_var[2], "%)"),
         title = "PCA of CLR profiles — diabetics: East vs. West") +
    theme_bw()
```

### Constrained Ordination — RDA

The CLR matrix is constrained by country, sex, age, and BMI. The term of interest
is `Country` (West relative to China), with sex, age, and BMI as covariates.

```{r a3-rda}
rda_we_out <- run_rda(
  ps_clr      = we_t$clr,
  formula_str = "Country + as.numeric(Host.age) + as.numeric(Nr..reads.sequenced)",
  covariates  = c("Country", "Host.age", "Nr..reads.sequenced")
)

anova(rda_we_out$rda, permutations = 999)
anova(rda_we_out$rda, by = "term", permutations = 999)
```

```{r a3-rda-biplot, fig.width=9, fig.height=6}
plot_rda_biplot(
  rda_obj      = rda_we_out$rda,
  meta_df      = rda_we_out$meta,
  color_var    = "Country",
  shape_var    = "Sex",
  color_values = c("China" = "#E74C3C", "Europe" = "#3498DB"),
  shape_values = c("Female" = 16, "Male" = 17),
  title        = "RDA biplot — diabetics: East vs. West"
)
```

### Compare RDA and CLR
```{r a3-rda-clr-compare, fig.width=8, fig.height=5}
rda1_scores_we <- scores(rda_we_out$rda, display = "sites")[, "RDA1"]
rda1_sign_we <- ifelse(
  mean(rda1_scores_we[rda_we_out$meta$Country == "Europe"],  na.rm = TRUE) >=
  mean(rda1_scores_we[rda_we_out$meta$Country == "China"], na.rm = TRUE),
  1, -1
)

comp_we <- compare_rda_clr(
  rda_obj  = rda_we_out$rda,
  clr_res  = clr_we,
  tax_df   = as(tax_table(we_t$clr), "matrix") %>% as.data.frame(),
  rda1_sign = rda1_sign_we
)

plot_rda_clr_comparison(
  comp_we,
  x_label = "CLR estimate (diabetic vs healthy)",
  y_label = "RDA1 loading (positive = diabetic side)",
  title   = "RDA loadings vs CLR estimates — global"
)
```

```{r a3-rda-clr-table}
comp_we %>%
  arrange((abs(clr_q))) %>%
  select(taxon, Genus.x, Species.x, rda1_loading, clr_estimate, clr_q, direction_agreement) %>%
  head(30)
```