---
title: "Multivariate_statistics"
output: html_document
---

```{r}
library(phyloseq)
library(vegan)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(ggplot2)
library(tidyverse)
library(ggh4x)
library(lme4)
library(lmerTest)
library(broom.mixed)
```

```{r}
path <- "../data/D003924_Diabetes_Mellitus,_T/species/phyloseq.RData"

ps <- get(load(path))

meta <- as(sample_data(ps), "matrix") %>% as.data.frame()
otu <- as(otu_table(ps), "matrix") %>% as.data.frame()
tax <- as(tax_table(ps), "matrix") %>% as.data.frame()

ps_rel <- transform_sample_counts(ps, function(x) x / sum(x))
otu_rel <- as(otu_table(ps_rel), "matrix") %>% t() %>% as.data.frame()
meta_rel <- as(sample_data(ps_rel), "matrix") %>% as.data.frame()
tax_rel <- as(tax_table(ps_rel), "matrix") %>% as.data.frame()  
```

# Compositional analysis with CLR transformation

Details: [9. Compositional Analysis with CLR Transformation][wiki-clr]. Transform counts with CLR before fitting the adjusted age model on a compositional scale.

## CLR transform
```{r}
ps_clr <- transform_sample_counts(ps, function(x) {
  x <- x + 1e-9
  log(x) - mean(log(x))
})
otu_clr <- as(otu_table(ps_clr), "matrix") %>% t() %>% as.data.frame()
meta_clr <- as(sample_data(ps_clr), "matrix") %>% as.data.frame()
tax_clr <- as(tax_table(ps_clr), "matrix") %>% as.data.frame()
```
The hidden chunk below reshapes CLR values into the long table `dat_clr`.

```{r, include=FALSE}
dat_clr <- otu_clr %>%
  rownames_to_column("sample_id") %>%
  pivot_longer(-sample_id, names_to = "taxon", values_to = "clr") %>%
  left_join(meta_clr %>% as("data.frame") %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  left_join(tax_clr %>% rownames_to_column("taxon"), by = "taxon")
```
Fit one adjusted linear model per taxon using CLR values as the response.

# Multivariate statistics

These sections analyse all gut taxa jointly on the CLR scale. PCA is unconstrained; RDA is constrained by metadata.

## PCA of CLR-transformed gut profiles

Details: [11. PCA of CLR-Transformed Gut Profiles][wiki-pca]. Build a gut-only CLR matrix and inspect the largest unconstrained community gradients.

```{r}
gut_clr_mat <- dat_clr %>%
  filter(body_site == "gut") %>%
  select(sample_id, taxon, clr) %>%
  pivot_wider(names_from = taxon, values_from = clr) %>%
  column_to_rownames("sample_id") %>%
  as.matrix()

gut_meta <- meta_clr[rownames(gut_clr_mat), ] %>% as("data.frame")

gut_pca <- prcomp(gut_clr_mat, center = TRUE, scale. = FALSE)

pca_scores <- as.data.frame(gut_pca$x[, 1:2]) %>%
  rownames_to_column("sample_id") %>%
  left_join(gut_meta %>% rownames_to_column("sample_id"), by = "sample_id")

pca_var <- round(100 * gut_pca$sdev^2 / sum(gut_pca$sdev^2), 1)

pca_scores %>%
  ggplot(aes(x = PC1, y = PC2, color = age, shape = status)) +
    geom_point(size = 3, alpha = 0.85) +
    scale_color_manual(values = c("adult" = "#E74C3C", "infant" = "#3498DB")) +
    scale_shape_manual(values = c("healthy" = 16, "sick" = 17)) +
    labs(x = paste0("PC1 (", pca_var[1], "%)"),
         y = paste0("PC2 (", pca_var[2], "%)"),
         color = "Age", shape = "Status",
         title = "PCA of gut microbiome CLR profiles") +
    theme_bw()
```
Use PCA to see whether age or health status align with the strongest overall community gradients.

## Constrained ordination with RDA

Details: [12. Constrained Ordination with RDA][wiki-rda]. Fit an RDA on the gut CLR matrix constrained by age and health status.

```{r}
rda_age <- rda(gut_clr_mat ~ age + status, data = gut_meta)
```
The sample order in `gut_meta` must match the rows of `gut_clr_mat`.

Use permutation ANOVA to test the constrained model and each predictor.

```{r}
anova(rda_age, permutations = 999)
anova(rda_age, by = "term", permutations = 999)
```
A significant age term supports a community-level age effect after accounting for status.

Extract RDA1 taxon scores to rank the taxa driving the primary constrained axis.

```{r}
rda_site_scores_raw <- scores(rda_age, display = "sites")
rda1_sign <- ifelse(
  mean(rda_site_scores_raw[gut_meta$age == "adult", "RDA1"], na.rm = TRUE) >=
    mean(rda_site_scores_raw[gut_meta$age == "infant", "RDA1"], na.rm = TRUE),
  1, -1
)

taxon_scores <- scores(rda_age, display = "species")[, 1] * rda1_sign

rda_taxa <- tibble(
  taxon = names(taxon_scores),
  rda1_loading = taxon_scores
) %>%
  arrange(desc(abs(rda1_loading)))
```
The `rda_taxa` table ranks taxa by absolute loading magnitude.

The biplot below shows samples, predictors and the top RDA1 taxa in the same constrained ordination space.

```{r}
eig <- rda_age$CCA$eig
pct1 <- round(eig[1] / rda_age$tot.chi * 100, 1)
pct2 <- round(eig[2] / rda_age$tot.chi * 100, 1)

site_scores <- scores(rda_age, display = "sites") %>%
  as.data.frame() %>%
  mutate(RDA1 = RDA1 * rda1_sign) %>%
  rownames_to_column("sample_id") %>%
  left_join(gut_meta %>% rownames_to_column("sample_id"), by = "sample_id")

biplot_scores <- scores(rda_age, display = "bp") %>%
  as.data.frame() %>%
  mutate(RDA1 = RDA1 * rda1_sign) %>%
  rownames_to_column("variable")

top_species_scores <- scores(rda_age, display = "species") %>%
  as.data.frame() %>%
  mutate(RDA1 = RDA1 * rda1_sign) %>%
  rownames_to_column("taxon") %>%
  inner_join(rda_taxa %>% head(10), by = "taxon") %>%
  left_join(tax_clr %>% rownames_to_column("taxon"), by = "taxon") %>%
  mutate(label = species)

# Scale arrows so the longest reaches ~80 % of the site score range
site_range <- max(abs(c(site_scores$RDA1, site_scores$RDA2)))
arrow_scale   <- site_range * 0.8 /
  max(sqrt(biplot_scores$RDA1^2   + biplot_scores$RDA2^2))
species_scale <- site_range * 0.6 /
  max(sqrt(top_species_scores$RDA1^2 + top_species_scores$RDA2^2))

biplot_scores      <- biplot_scores      %>% mutate(RDA1 = RDA1 * arrow_scale,   RDA2 = RDA2 * arrow_scale)
top_species_scores <- top_species_scores %>% mutate(RDA1 = RDA1 * species_scale, RDA2 = RDA2 * species_scale)

ggplot() +
  geom_point(data = site_scores,
             aes(x = RDA1, y = RDA2, color = age, shape = status),
             size = 3, alpha = 0.85) +
  geom_segment(data = top_species_scores,
               aes(x = 0, y = 0, xend = RDA1, yend = RDA2),
               arrow = arrow(length = unit(0.15, "cm")),
               color = "grey70", linewidth = 0.5) +
  geom_text(data = top_species_scores,
            aes(x = RDA1 * 1.15, y = RDA2 * 1.15, label = label),
            size = 2.5, color = "grey50") +
  geom_segment(data = biplot_scores,
               aes(x = 0, y = 0, xend = RDA1, yend = RDA2),
               arrow = arrow(length = unit(0.25, "cm")),
               color = "black", linewidth = 1) +
  geom_text(data = biplot_scores,
            aes(x = RDA1 * 1.15, y = RDA2 * 1.15, label = variable),
            size = 3.5, fontface = "bold", color = "black") +
  scale_color_manual(values = c("adult" = "#E74C3C", "infant" = "#3498DB")) +
  scale_shape_manual(values = c("healthy" = 16, "sick" = 17)) +
  labs(x = paste0("RDA1 (", pct1, "% of total variance)"),
       y = paste0("RDA2 (", pct2, "% of total variance)"),
       color = "Age", shape = "Status",
       title = "RDA biplot: gut microbiome constrained by age and health status") +
  theme_bw()
```
Look for age separation among sample points and for taxa arrows aligned with that gradient.

The final plot ranks the strongest RDA1 taxon loadings directly.

```{r}
rda_taxa %>%
  head(20) %>%
  left_join(tax_clr %>% rownames_to_column("taxon"), by = "taxon") %>%
  mutate(
    label = species,
    direction = ifelse(rda1_loading > 0, "Adult-associated", "Infant-associated")
  ) %>%
  ggplot(aes(x = reorder(label, rda1_loading), y = rda1_loading, fill = direction)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("Adult-associated" = "#E74C3C", "Infant-associated" = "#3498DB")) +
    labs(x = NULL, y = "RDA1 loading", fill = NULL,
         title = "Top 20 taxa by RDA1 loading magnitude") +
    theme_bw()
```
Compare these RDA drivers with the consensus taxa from the univariate methods.

The final comparison links the multivariate RDA loadings to the CLR-based univariate age estimates.

```{r}
rda_clr_comparison <- rda_taxa %>%
  left_join(
    clr_age_adjusted_status_results %>%
      select(taxon, genus, species,
             clr_estimate = estimate,
             clr_q = q_value),
    by = "taxon"
  ) %>%
  mutate(
    direction_agreement = sign(rda1_loading) == sign(clr_estimate),
    clr_result = case_when(
      clr_q < 0.05 & direction_agreement ~ "CLR significant, same direction",
      clr_q < 0.05                      ~ "CLR significant, opposite direction",
      TRUE                              ~ "CLR not significant"
    )
  )

rda_clr_comparison %>%
  ggplot(aes(x = clr_estimate, y = rda1_loading, color = clr_result)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    geom_point(alpha = 0.65, size = 2) +
    scale_color_manual(values = c(
      "CLR significant, same direction"     = "#2E8B57",
      "CLR significant, opposite direction" = "#E67E22",
      "CLR not significant"                 = "grey70"
    )) +
    labs(x = "CLR age estimate (adult vs infant)",
         y = "RDA1 loading (positive = adult side)",
         color = NULL,
         title = "RDA loadings vs CLR univariate age effects") +
    theme_bw()
```
Points in the upper-right and lower-left quadrants agree in direction between RDA and CLR.

```{r}
rda_clr_comparison %>%
  arrange(desc(abs(rda1_loading))) %>%
  select(genus, species, rda1_loading, clr_estimate, clr_q, direction_agreement) %>%
  head(20)
```
Use [13. How to Read the Practical Results][wiki-read-results] for the final interpretation.