---
title: "CLR_linearmodel"
output: html_document
---

```{r}
library(phyloseq)
library(vegan)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(ggplot2)
library(ape)
library(tidyverse)
library(ggh4x)
library(lme4)
library(lmerTest)
library(broom.mixed)
```

# Explore and load data
## Load data
```{r}
path <- "../data/molecular_ecology_17695.rds"

phyloseqobj <- read_rds(path)$phyloseq

meta <- as(sample_data(phyloseqobj), "matrix") %>% as.data.frame()
otu <- as(otu_table(phyloseqobj), "matrix") %>% as.data.frame()
tax <- as(tax_table(phyloseqobj), "matrix") %>% as.data.frame()
tree <- phy_tree(phyloseqobj)

```
## Counts
```{r}
table(meta$origin, meta$location, useNA = "always")

max(sample_sums(phyloseqobj))


```
## Cleaning data
```{r build-phyloseq, include=FALSE}
ps <- phyloseqobj

# Note: data is relative abundance - Observed/Chao1 not valid
ps_rel <- transform_sample_counts(ps, function(x) x / sum(x))
ps_pa  <- transform_sample_counts(ps, function(x) ifelse(x > 0, 1, 0))

```


## Barplot for sick and healthy
### Sample function
```{r}
sample_ps <- function(ps, FUN = sample, ...){
  ids <- sample_names(ps)
  sampled_ids <- FUN(ids, ...)
  ps <- prune_samples(sampled_ids, ps)
  return(ps)
}
```

### Barplot
```{r}
ps_rel <- transform_sample_counts(ps, function(x) x / sum(x))
otu_rel <- as(otu_table(ps_rel), "matrix") %>% t() %>% as.data.frame()
meta_rel <- as(sample_data(ps_rel), "matrix") %>% as.data.frame()
tax_rel <- as(tax_table(ps_rel), "matrix") %>% as.data.frame()  
ps_genus <- tax_glom(ps_rel, taxrank = "genus")

ps_genus_sampled <- sample_ps(ps_genus, FUN = sample, size = 40)

plot_bar(ps_genus, fill = "genus") +
  facet_nested(~ location, scales = "free_x") +
  theme(axis.text.x = element_blank())
```

## Tree
```{r}
plot_tree(ps, method = "treeonly", ladderize = TRUE) 
```
## Heatmap
```{r}
plot_heatmap(ps_genus_sampled, taxa.label = "species")
```




# Alpha diversity
## Plot Shannon index and Observed abundances
**Must use untrimmed data for meaningful results. Ours is not**
```{r}
plot_richness(
  ps,
  measures = "Shannon",
  color = "location"
)

ggplot(alpha_diversity, aes(x = Disease.MESH.ID, y = Shannon)) +
  geom_boxplot() +
  #geom_jitter(width = 0.2) +
  theme_bw()

plot_richness(
  ps_pa,
  measures = "Observed",
  color = 'location'
)
```
## Statistical testing on alpha diversity
```{r}
ps_pa <- transform_sample_counts(ps, function(x) {ifelse(x > 0, 1, 0)})

# Calculate Shannon diversity
shannon <- estimate_richness(
  ps,
  measures = "Shannon"
) %>%
  rownames_to_column("sample_id")

# Calculate observed richness
observed <- estimate_richness(
  ps_pa,
  measures = "Observed"
) %>%
  rownames_to_column("sample_id")

# Extract sample metadata
sample_df <- sample_data(ps) %>%
  data.frame() %>%
  rownames_to_column("sample_id")

# Combine diversity estimates and metadata
alpha_diversity <- inner_join(
  shannon,
  observed,
  by = "sample_id"
) %>%
  inner_join(
    sample_df,
    by = "sample_id"
  )

head(alpha_diversity)

```
### Kruskal
```{r}
# Create a function to run multiple tests
run_kruskal_tests <- function(data, response_var, group_vars) {
  results <- list()
  
  for (group_var in group_vars) {
    formula_obj <- as.formula(paste(response_var, "~", group_var))
    test_result <- kruskal.test(formula_obj, data = data)
    
    results[[group_var]] <- data.frame(
      Variable = group_var,
      Chi_squared = test_result$statistic,
      df = test_result$parameter,
      P_value = test_result$p.value
    )
  }
  
  return(do.call(rbind, results))
}

# Use the function
kruskal_table <- run_kruskal_tests(
  data = alpha_diversity,
  response_var = "Shannon",
  group_vars = c("origin", "location")
)

# View results
print(kruskal_table)
```
### Pairwise wilcoxon
```{r pairwise-wilcoxon-shannon}
run_pairwise_wilcoxon <- function(data, response_var, group_vars) {
  results <- list()
  
  for (group_var in group_vars) {
    formula_obj <- as.formula(paste(response_var, "~", group_var))
    
    # Run pairwise Wilcoxon test
    test_result <- pairwise.wilcox.test(
      data[[response_var]], 
      data[[group_var]],
      p.adjust.method = "bonferroni"  # Can change to "holm", "fdr", etc.
    )
    
    # Convert the result matrix to a tidy data frame
    p_values <- as.data.frame(test_result$p.value)
    p_values$Group1 <- rownames(p_values)
    
    # Reshape to long format
    p_values_long <- p_values %>%
      pivot_longer(cols = -Group1, 
                   names_to = "Group2", 
                   values_to = "P_value") %>%
      drop_na(P_value) %>%
      mutate(
        Variable = group_var,
        Comparison = paste(Group1, "vs", Group2),
        P_value_adj = P_value
      ) %>%
      select(Variable, Comparison, P_value_adj, Group1, Group2)
    
    results[[group_var]] <- p_values_long
  }
  
  return(do.call(rbind, results))
}

# Run the function
pairwise_wilcox_results <- run_pairwise_wilcoxon(
  data = alpha_diversity,
  response_var = "Shannon",
  group_vars = c("origin", "location")
)

# View results
print(pairwise_wilcox_results)
```

# Depth correlated with sequencing depth
```{r depth-vs-shannon, fig.width=7, fig.height=5}
# Total abundance per sample
sample_depth <- sample_sums(ps)

alpha_diversity$reads <- sample_depth[
  alpha_diversity$sample_id
]

ggplot(alpha_diversity, aes(x = reads, y = Shannon)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_bw()

cor.test(
  alpha_diversity$reads,
  alpha_diversity$Shannon,
  method = "spearman"
)
```
# Beta-diversity
## Bray Curtis
```{r calculate-bray-distance}
bray_dist <- distance(
  ps_rel,
  method = "bray"
)

weighted_unifrac <- distance(
  ps_rel,
  method = "wunifrac"
)

jaccard_dist <- distance(
  ps_pa,
  method = "jaccard"
)
unifrac_dist <- distance(
  ps_pa,
  method = "unifrac"
)
```

# Ordination
## PCoA based on Bray-Curtis distance

PCoA is a useful first ordination method because it attempts to represent the pairwise sample distances in a lower-dimensional Euclidean space.

```{r pcoa-bray-basic, fig.width=8, fig.height=5}
ord_bray <- ordinate(
  ps_rel,
  method = "PCoA",
  distance = bray_dist
)

plot_ordination(
  ps_rel,
  ord_bray,
  type = "samples",
  color = "location",
  shape = 'origin'
)
```

The resulting plot can be used to explore whether samples from the same body site tend to cluster together and whether different body-site categories appear compositionally distinct.

## Visualising the explained variation

For PCoA, it is often useful to inspect how much of the variation is represented by the first few axes.

```{r pcoa-scree-plot, fig.width=8, fig.height=5}
plot_scree(ord_bray)
```

## Extracting ordination coordinates

The built-in plot is useful for quick inspection, but it is often convenient to extract the ordination coordinates as a standard data frame so that we can customise the visualisation further with ggplot2.

```{r build-ordination-data-frame}
ord_bray_df <- plot_ordination(
  ps_rel,
  ord_bray,
  type = "samples",
  justDF = TRUE
)

head(ord_bray_df)
```

Once the coordinates and metadata are into a single table, we can create a more flexible plot:

```{r pcoa-bray-ggplot, fig.width=8, fig.height=5}
ggplot(ord_bray_df, aes(x = Axis.1, y = Axis.2, color = location)) +
  geom_point(size = 3) +
  theme_bw()
```

## NMDS based on Bray-Curtis distance

NMDS is another commonly used ordination method in ecology and microbiome analysis. Unlike PCoA, it focuses on preserving the rank order of distances rather than the distances themselves.

```{r nmds-bray-basic, fig.width=8, fig.height=5}
ord_nmds <- ordinate(
  ps_rel,
  method = "NMDS",
  distance = "bray"
)

plot_ordination(
  ps_rel,
  ord_nmds,
  type = "samples",
  color = "location"
)
```

NMDS is often particularly useful when ecological relationships are non-linear, although the results can vary slightly between runs because the algorithm uses an iterative optimisation procedure.

## Statistical testing - betadiversity

Ordination plots are very useful for quick exploratory visualisations, but in many analyses we also want to perform formal statistical testing to evaluate whether community composition differs significantly between groups of samples.

For this, we will use PERMANOVA (Permutational Multivariate Analysis of Variance), implemented in the `vegan` package through the function `adonis2()`.

### Combining beta-diversity distances with sample metadata

Before running the test, it is convenient to extract the sample metadata stored in the phyloseq object.

```{r extract-sample-metadata}
sample_df <- sample_data(ps_rel) %>%
  data.frame()

head(sample_df)

adonis_bray <- adonis2(
  bray_dist ~ location,
  data = sample_df,
  permutations = 999
)

adonis_bray
```


## Checking homogeneity of dispersion

One important limitation of PERMANOVA is that significant results can sometimes reflect not only differences in group centroids, but also differences in within-group dispersion.

We can explore this by using `betadisper()`.

```{r dispersion-test}
dispersion_bray <- betadisper(
  bray_dist,
  sample_df$origin
)

anova(dispersion_bray)
plot(dispersion_bray)
```

If dispersion differs strongly between groups, then a significant PERMANOVA result should be interpreted with caution.

## Visualising ordination with group structure

Statistical tests are much easier to interpret when combined with visualisations.

A common approach is to display the ordination plot together with group labels and inspect whether the observed separation is visually consistent with the statistical result.

```{r pcoa-bray-with-groups, fig.width=8, fig.height=5}
ggplot(ord_bray_df, aes(x = Axis.1, y = Axis.2, color = location)) +
  geom_point(size = 3) +
  theme_bw()
```

## Exercise: Comparing a subset of samples

In some cases, it may be useful to restrict the analysis to a subset of environments and ask whether the same broad pattern still holds.

For example, you may want to compare only the `gut_adult` and `gut_infant` groups.

```{r permanova-subset}
ps_rel_china <- subset_samples(
  ps_rel,
  Country %in% c("China")
)

bray_dist_china <- distance(
  ps_rel_china,
  method = "bray"
)

sample_df_china <- sample_data(ps_rel_china) %>%
  data.frame()

adonis2(
  bray_dist_china ~ Disease.MESH.ID,
  data = sample_df_china,
  permutations = 999
)
```


# Univariate statistics: Compositional analysis with CLR transformation

Details: [9. Compositional Analysis with CLR Transformation][wiki-clr]. Transform counts with CLR before fitting the adjusted age model on a compositional scale.

## CLR
```{r}
ps_clr <- transform_sample_counts(ps, function(x) {
  x <- x + 1e-9
  log(x) - mean(log(x))
})
otu_clr <- as(otu_table(ps_clr), "matrix") %>% t() %>% as.data.frame()
meta_clr <- as(sample_data(ps_clr), "matrix") %>% as.data.frame()
tax_clr <- as(tax_table(ps_clr), "matrix") %>% as.data.frame()

```
The hidden chunk below reshapes CLR values into the long table `dat_clr`.

```{r, include=FALSE}
dat_clr <- otu_clr %>%
  rownames_to_column("sample_id") %>%
  pivot_longer(-sample_id, names_to = "taxon", values_to = "clr") %>%
  left_join(meta_clr %>% as("data.frame") %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  left_join(tax_clr %>% rownames_to_column("taxon"), by = "taxon")
```
Fit one adjusted linear model per taxon using CLR values as the response.

## CLR
```{r}
clr_status <- dat_clr %>%
  mutate(
    origin = factor(origin, levels = c("Feral", "Domestic")),
    location = factor(location),
  ) %>%
  #filter(!is.na(age), !is.na(bmi)) %>%
  group_by(taxon, genus, species) %>%
  group_modify(~ {
    fit <- lm(
      clr ~ origin + location,
      data = .x
    )

    broom::tidy(fit)
  }) %>%
  filter(term == 'originDomestic') %>%
  ungroup() %>%
  mutate(q_value = p.adjust(p.value, method = "BH"))
```
Positive CLR estimates indicate adult enrichment relative to the community geometric mean; negative estimates indicate infant enrichment. The next table lists infant-enriched taxa.

```{r}
clr_healthy_enriched <- clr_status %>%
  select(taxon, genus, species, estimate, std.error, statistic, p.value, q_value) %>%
  arrange(estimate) %>%
  head(10)

clr_healthy_enriched
```


```{r}
dat_clr %>%
  filter(taxon == clr_healthy_enriched %>% slice(1) %>% pull(taxon)) %>%
  ggplot(aes(x = origin, y = clr, fill = origin)) +
    geom_boxplot(outlier.shape = NA) +
    #geom_jitter(width = 0.15) +
    theme_bw()
```
The y-axis is CLR, not relative abundance; compare direction rather than raw magnitude with earlier plots.

The next table lists CLR adult-enriched taxa.

```{r}
clr_sick_enriched <- clr_status %>%
  select(taxon, genus, species, estimate, std.error, statistic, p.value, q_value) %>%
  arrange(-estimate) %>%
  head(10)

clr_sick_enriched
```
Inspect the top CLR adult-enriched taxon on the CLR scale.

```{r}
dat_clr %>%
  filter(taxon == clr_sick_enriched %>% slice(1) %>% pull(taxon)) %>%
  ggplot(aes(x = origin, y = clr, fill = origin)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15) +
    theme_bw()
```
Check whether the group separation matches the CLR estimate and q-value.

The CLR volcano plot summarises the compositional model results.

```{r}
# Volcano plot: CLR-based linear model
clr_status %>%
  mutate(
    neg_log10_q = -log10(q_value),
    direction = case_when(
      q_value < 0.05 & estimate > 0 ~ "Sick-enriched",
      q_value < 0.05 & estimate < 0 ~ "Healthy-enriched",
      TRUE ~ "Not significant"
    )
  ) %>%
  ggplot(aes(x = estimate, y = neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 2) +
    scale_color_manual(values = c("Sick-enriched" = "#E74C3C", "Healthy-enriched" = "#3498DB", "Not significant" = "grey60")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    labs(x = "CLR estimate (adult vs infant)", y = "-log10(q-value)", color = NULL,
         title = "Compositional analysis (CLR): age effect in gut microbiome") +
    theme_bw()
```
Use this volcano plot as the compositionally aware reference analysis.

The next plot shows CLR effect size on the x-axis and orders taxa by q-value.

```{r}
# Bar chart of top significant taxa ordered by q-value (CLR)
bind_rows(
  clr_status %>% filter(q_value < 0.05) %>% arrange(estimate) %>% head(10),
  clr_status %>% filter(q_value < 0.05) %>% arrange(-estimate) %>% head(10)
) %>%
  distinct() %>%
  mutate(
    label = species,
    direction = ifelse(estimate > 0, "Female-enriched", "Male-enriched")
  ) %>%
  ggplot(aes(x = reorder(label, -q_value), y = estimate, fill = direction)) +
    geom_col() +
    coord_flip() +
    scale_color_manual(values = c("Female-enriched" = "#E74C3C", "Male-enriched" = "#3498DB")) +
    labs(x = NULL, y = "CLR estimate (Female vs male)", fill = NULL,
         title = "Top differentially abundant taxa (CLR)") +
    theme_bw()
```
Taxa that change strongly under CLR are important interpretation cases.

# Multivariate statistics
## PCA of CLR-transformed gut profiles

Details: [11. PCA of CLR-Transformed Gut Profiles][wiki-pca]. Build a gut-only CLR matrix and inspect the largest unconstrained community gradients.

```{r}
clr_mat <- dat_clr %>%
  select(sample_id, taxon, clr) %>%
  pivot_wider(names_from = taxon, values_from = clr) %>%
  column_to_rownames("sample_id") %>%
  as.matrix()

meta_clr_df <- meta_clr[rownames(clr_mat), ] %>% as("data.frame")

clr_pca <- prcomp(clr_mat, center = TRUE, scale. = FALSE)

pca_scores <- as.data.frame(clr_pca$x[, 1:2]) %>%
  rownames_to_column("sample_id") %>%
  left_join(meta_clr_df %>% rownames_to_column("sample_id"), by = "sample_id")

pca_var <- round(100 * clr_pca$sdev^2 / sum(clr_pca$sdev^2), 1)

pca_scores %>%
  ggplot(aes(x = PC1, y = PC2, color = factor(origin), shape = factor(location))) +
    geom_point(size = 3, alpha = 0.85) +
    scale_color_manual(values = c("Feral" = "#E74C3C", "Domestic" = "#3498DB")) +
    scale_shape_manual(values = c("Aruba" = 16, "Brazil" = 17, "CaboVerde" = 18, "Denmark" = 19, "Malaysia" = 20, "Spain" = 21)) +
    labs(x = paste0("PC1 (", pca_var[1], "%)"),
         y = paste0("PC2 (", pca_var[2], "%)"),
         color = "Age", shape = "Status",
         title = "PCA of gut microbiome CLR profiles") +
    theme_bw()
```
Use PCA to see whether age or health status align with the strongest overall community gradients.

## Constrained ordination with RDA

Details: [12. Constrained Ordination with RDA][wiki-rda]. Fit an RDA on the gut CLR matrix constrained by age and health status.

```{r}
# Find which columns have NAs and how many
na_summary <- meta_clr_df %>%
  summarise(across(c(origin, location), 
                   ~ sum(is.na(.))))

print(na_summary)

# Find which samples have NAs
samples_with_na <- meta_clr_df %>%
  filter(if_any(c(origin, location), is.na)) %>%
  rownames()

print(paste("Samples with NAs:", length(samples_with_na)))
print(samples_with_na)

# Remove just those samples
meta_clr_clean <- meta_clr_df[!rownames(meta_clr_df) %in% samples_with_na, ]
clr_mat_clean <- clr_mat[!rownames(clr_mat) %in% samples_with_na, ]

rda_age <- rda(clr_mat_clean ~ origin + location, 
               data = meta_clr_clean)               
```

The sample order in `gut_meta` must match the rows of `gut_clr_mat`.

Use permutation ANOVA to test the constrained model and each predictor.

```{r}
anova(rda_age, permutations = 999)
anova(rda_age, by = "term", permutations = 999)
```

```{r}
rda_site_scores_raw <- scores(rda_age, display = "sites")
rda1_sign <- ifelse(
  mean(rda_site_scores_raw[meta_clr_clean$origin == "Feral", "RDA1"], na.rm = TRUE) >=
    mean(rda_site_scores_raw[meta_clr_clean$origin == "Domestic", "RDA1"], na.rm = TRUE),
  1, -1
)

taxon_scores <- scores(rda_age, display = "species")[, 1] * rda1_sign

rda_taxa <- tibble(
  taxon = names(taxon_scores),
  rda1_loading = taxon_scores
) %>%
  arrange(desc(abs(rda1_loading)))
```
The `rda_taxa` table ranks taxa by absolute loading magnitude.

The biplot below shows samples, predictors and the top RDA1 taxa in the same constrained ordination space.

```{r}
eig <- rda_age$CCA$eig
pct1 <- round(eig[1] / rda_age$tot.chi * 100, 1)
pct2 <- round(eig[2] / rda_age$tot.chi * 100, 1)

site_scores <- scores(rda_age, display = "sites") %>%
  as.data.frame() %>%
  mutate(RDA1 = RDA1 * rda1_sign) %>%
  rownames_to_column("sample_id") %>%
  left_join(meta_clr_clean %>% rownames_to_column("sample_id"), by = "sample_id")

biplot_scores <- scores(rda_age, display = "bp") %>%
  as.data.frame() %>%
  mutate(RDA1 = RDA1 * rda1_sign) %>%
  rownames_to_column("variable")

top_species_scores <- scores(rda_age, display = "species") %>%
  as.data.frame() %>%
  mutate(RDA1 = RDA1 * rda1_sign) %>%
  rownames_to_column("taxon") %>%
  inner_join(rda_taxa %>% head(10), by = "taxon") %>%
  left_join(tax_clr %>% rownames_to_column("taxon"), by = "taxon") %>%
  mutate(label = species)

# Scale arrows so the longest reaches ~80 % of the site score range
site_range <- max(abs(c(site_scores$RDA1, site_scores$RDA2)))
arrow_scale   <- site_range * 0.8 /
  max(sqrt(biplot_scores$RDA1^2   + biplot_scores$RDA2^2))
species_scale <- site_range * 0.6 /
  max(sqrt(top_species_scores$RDA1^2 + top_species_scores$RDA2^2))

biplot_scores      <- biplot_scores      %>% mutate(RDA1 = RDA1 * arrow_scale,   RDA2 = RDA2 * arrow_scale)
top_species_scores <- top_species_scores %>% mutate(RDA1 = RDA1 * species_scale, RDA2 = RDA2 * species_scale)

ggplot() +
  geom_point(data = site_scores,
             aes(x = RDA1, y = RDA2, color = origin, shape = location),
             size = 3, alpha = 0.85) +
  geom_segment(data = top_species_scores,
               aes(x = 0, y = 0, xend = RDA1, yend = RDA2),
               arrow = arrow(length = unit(0.15, "cm")),
               color = "grey70", linewidth = 0.5) +
  geom_text(data = top_species_scores,
            aes(x = RDA1 * 1.15, y = RDA2 * 1.15, label = label),
            size = 2.5, color = "grey50") +
  geom_segment(data = biplot_scores,
               aes(x = 0, y = 0, xend = RDA1, yend = RDA2),
               arrow = arrow(length = unit(0.25, "cm")),
               color = "black", linewidth = 1) +
  geom_text(data = biplot_scores,
            aes(x = RDA1 * 1.15, y = RDA2 * 1.15, label = variable),
            size = 3.5, fontface = "bold", color = "black") +
    scale_color_manual(values = c("Feral" = "#E74C3C", "Domestic" = "#3498DB")) +
    scale_shape_manual(values = c("Aruba" = 16, "Brazil" = 17, "CaboVerde" = 18, "Denmark" = 19, "Malaysia" = 20, "Spain" = 21)) 
  labs(x = paste0("RDA1 (", pct1, "% of total variance)"),
       y = paste0("RDA2 (", pct2, "% of total variance)"),
       color = "Age", shape = "Status",
       title = "RDA biplot: gut microbiome constrained by age and health status") +
  theme_bw()
```
Look for age separation among sample points and for taxa arrows aligned with that gradient.

The final plot ranks the strongest RDA1 taxon loadings directly.

```{r}
rda_taxa %>%
  head(20) %>%
  left_join(tax_clr %>% rownames_to_column("taxon"), by = "taxon") %>%
  mutate(
    label = species,
    direction = ifelse(rda1_loading > 0, "Adult-associated", "Infant-associated")
  ) %>%
  ggplot(aes(x = reorder(label, rda1_loading), y = rda1_loading, fill = direction)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("Adult-associated" = "#E74C3C", "Infant-associated" = "#3498DB")) +
    labs(x = NULL, y = "RDA1 loading", fill = NULL,
         title = "Top 20 taxa by RDA1 loading magnitude") +
    theme_bw()
```
Compare these RDA drivers with the consensus taxa from the univariate methods.

The final comparison links the multivariate RDA loadings to the CLR-based univariate age estimates.

```{r}
rda_clr_comparison <- rda_taxa %>%
  left_join(
    clr_status %>%
      select(taxon, genus, species,
             clr_estimate = estimate,
             clr_q = q_value),
    by = "taxon"
  ) %>%
  mutate(
    direction_agreement = sign(rda1_loading) == sign(clr_estimate),
    clr_result = case_when(
      clr_q < 0.05 & direction_agreement ~ "CLR significant, same direction",
      clr_q < 0.05                      ~ "CLR significant, opposite direction",
      TRUE                              ~ "CLR not significant"
    )
  )

rda_clr_comparison %>%
  ggplot(aes(x = clr_estimate, y = rda1_loading, color = clr_result)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    geom_point(alpha = 0.65, size = 2) +
    scale_color_manual(values = c(
      "CLR significant, same direction"     = "#2E8B57",
      "CLR significant, opposite direction" = "#E67E22",
      "CLR not significant"                 = "grey70"
    )) +
    labs(x = "CLR age estimate (adult vs infant)",
         y = "RDA1 loading (positive = adult side)",
         color = NULL,
         title = "RDA loadings vs CLR univariate age effects") +
    theme_bw()
```
Points in the upper-right and lower-left quadrants agree in direction between RDA and CLR.

```{r}
rda_clr_comparison %>%
  arrange(desc(abs(rda1_loading))) %>%
  select(genus, species, rda1_loading, clr_estimate, clr_q, direction_agreement) %>%
  head(20)
```
Use [13. How to Read the Practical Results][wiki-read-results] for the final interpretation.

